Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-11 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
edits in 5.8

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,3)$ and $\mathrm{Gamma}(1.2,0.2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variance $\alpha$, and length-scale $\ell$, respectively. We use the same prior distributions to fit the exact GP model as the HSGP models.

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,3)$ and $\mathrm{InverseGamma}(2,0.5)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variance $\alpha$, and length-scales $\ell$, respectively. We use the same prior distributions to fit the exact GP model as the HSGP models on the synthetic data sets.

The HSGP models are fitted following the recommended iterative steps as explained in Section~\ref{sec_user_guide}, where in each iteration the diagnosis is applied on every dimensions and, for each one of the dimensions, the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated by equations in Section~\ref{sec_num_equations}. The values $0.5$, $1$, and $1$ are set as initial values for the minimum lengthscales $\ell_1$, $\ell_2$ and $\ell_3$, for the first, second and third dimensions, respectively. In order to be as more efficient as possible in these multivariate cases, after the diagnostic generated the first \textit{true} for a certain dimension, its corresponding $m$ is updated by increasing the $m$ of the previous iteration by only 2 additional basis functions and, after the second \textit{true}, $m$ is no longer increased. The full diagnosis process is applied until two \textit{trues} are achieved for each dimension.

Table~\ref{tab_caseIII_2D} and \ref{tab_caseIII_3D} contain the iterative steps to fit and diagnose the $2$D and $3$D data sets, respectively. 
%These tables contain the values for the parameters $\ell$, $c$, $m$, the estimated $\hat{\ell}$, the diagnostic $\hat{\ell} + 0.01 \geq \ell$ for each dimension, and the RMSE compute with both the data and the GP as the reference for every iterative steps of the fitting process. 
The minimum requirements to fit the models were easily achieved by performing 4 iterations: from a $1^{st}$ iteration that uses few multivariate basis functions $m^*$ ($m^*=m_1(=6) \times m_2(=6)=36$ and $m^*=m_1(=6) \times m_2(=6) \times m_3(=6)=216$ for the $2$D and $3$D data sets, respectively) to the $4^{th}$ iteration that uses the minimum multivariate basis functions ($m^*=m_1(=22) \times m_2(=11)=242$ and $m^*=m_1(=21) \times m_2(=10) \times m_3(=12)=2520$ for the $2$D and $3$D data sets, respectively) required to accurately fit these notably wiggly data sets. Without using our recommended guidelines and diagnostics, it could have taken user several trials to get the optimal solution and to expend a significant amount of time, as using a larger number of basis functions in multivariate cases increases time of computation drastically.

@@

\subsubsection{Computation requirements in $2$D, $3$D and $4$D input spaces}\label{sec_computation_nD}

Experiments have been conducted to empirically recognize time of computation of the HSGP model in multivariate cases and different number of sample data points. Dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} was used for the experiments. Implementation and sampling has been done only for the sample data.

Figure~\ref{fig_time_nD_log} shows the computation time (in seconds per iteration) in $2$D, $3$D and $4$D input spaces and different sized data sets, $n=300, n=500, n=1000$ and $n=3000$, and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation. Both the time of computation and $m^*$ are plotted in the logarithm scale.

If we look back at Figure~\ref{fig6_relationships} or equations~\eqref{eq_m_l_QE}, \eqref{eq_m_l_mat32} and \eqref{eq_m_l_mat52} we can find out that, for example for a squared exponential kernel, any univariate function (or single dimensions of a multivariate function) with true lengthscales bigger than $0.3$ can be accurately fitted, after applying the iterative diagnostic steps, by only using approximately $10$ or $12$ basis functions as much, with the assumption that we have chosen the optimal value of the boundary factor $c$. And, for true lengthscales between $0.1$ and $0.3$, by only using between $10$ and $22$ basis functions as much, with the assumption that we have chosen the optimal value of the boundary factor $c$.

So, for a very wiggly $2$D function (let's say with $\ell_1=0.1$ and $\ell_2=0.1$), the approximate number of multivariate basis functions needed for an accurate approximation will be $m*=22\times 22= 484$, which results in an approximation significantly faster than the regular GP even in very small sized data sets (i.e., $n<300$) (see Figure~\ref{fig_time_nD_log}). Similarly, for a very wiggly $3$D function (let's say with $\ell_1=0.1$, $\ell_2=0.1$ and $\ell_3=0.1$), we can expect an approximate minimum number of multivariate basis functions for an accurate approximation of $m*=22\times 22\times 22= 10648$, which, unless it is a small sized data set (i.e., $n<1000$), the method is significantly faster than the regular GP. Note the a $3$D function where each dimension has a lengthscale around $0.1$ is quite uncommon, so for most of the $3$D functions, unless it is a very small sized data set, the approximation will be significantly faster than the excat GP, if our iterative diagnostic rules are used as they allow to find out easily the solution with the minimum requirements. For $4$D data sets, the method can still be more efficient than the regular GP for moderate-large data sets (i.e., $n>1000$).

Finally, for $D>5$ the method starts to be impractical even for smooth univariate functions and small $n$. However, in these cases the method may still be used in an additive modeling scheme with low-dimensional additive components.

>>>>

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,3)$ and $\mathrm{Gamma}(1.2,0.2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variance $\alpha$, and length-scale $\ell$, respectively. We use the same prior distributions to fit the exact GP model.

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,3)$ and $\mathrm{InverseGamma}(2,0.5)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variance $\alpha$, and length-scales $\ell$, respectively. We use the same prior distributions to fit the exact GP model.

The HSGP models are fitted following the recommended iterative steps as explained in Section~\ref{sec_user_guide}, where in each iteration the diagnosis is applied on every dimensions and, for each one of the dimensions, the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated by equations in Section~\ref{sec_num_equations}. The values $0.5$, $1$, and $1$ are set as initial values for the minimum lengthscales $\ell_1$, $\ell_2$ and $\ell_3$, for the first, second and third dimensions, respectively. In order to be as efficient as possible in these multivariate cases, after the diagnostic generated the first \textit{true} for a certain dimension, its corresponding $m$ is updated by increasing the $m$ of the previous iteration by only 2 additional basis functions and, after the second \textit{true}, $m$ is no longer increased. The full diagnosis process is applied until two \textit{trues} are achieved for each dimension.

Table~\ref{tab_caseIII_2D} and \ref{tab_caseIII_3D} contain the iterative steps to fit and diagnose the $2$D and $3$D data sets, respectively. 
%These tables contain the values for the parameters $\ell$, $c$, $m$, the estimated $\hat{\ell}$, the diagnostic $\hat{\ell} + 0.01 \geq \ell$ for each dimension, and the RMSE compute with both the data and the GP as the reference for every iterative steps of the fitting process. 
The minimum requirements to fit the models were easily achieved by performing 4 iterations: from a $1^{st}$ iteration that uses few multivariate basis functions $m^*$ ($m^*=m_1(=6) \times m_2(=6)=36$ and $m^*=m_1(=6) \times m_2(=6) \times m_3(=6)=216$ for the $2$D and $3$D data sets, respectively) to the $4^{th}$ iteration that uses the minimum multivariate basis functions ($m^*=m_1(=22) \times m_2(=11)=242$ and $m^*=m_1(=21) \times m_2(=10) \times m_3(=12)=2520$ for the $2$D and $3$D data sets, respectively) required to accurately fit these notably wiggly data sets. Without using our recommended guidelines and diagnostics, it could have taken user several trials to get the optimal solution and to spend a significant amount of time, as using a larger number of basis functions in multivariate cases increases time of computation drastically.

@@

\subsubsection{Computation requirements in 2D, 3D and 4D input spaces}\label{sec_computation_nD}

%Experiments have been conducted to empirically recognize time of computation of the HSGP model in multivariate cases and different number of sample data points. Dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} was used for the experiments. Implementation and sampling has been done only for the sample data.

Figure~\ref{fig_time_nD_log} shows the computation time for dynamic HMC (in seconds per iteration) in $2$D, $3$D and $4$D input spaces and different sized data sets, $n=300, n=500, n=1000$ and $n=3000$, and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.
Both the time of computation and $m^*$ are plotted in the logarithm scale.

Looking back at Figure~\ref{fig6_relationships}, or equations~\eqref{eq_m_l_QE}, \eqref{eq_m_l_mat32} and \eqref{eq_m_l_mat52}, and assuming squared exponential covariance function, any univariate function (or single dimensions of a multivariate function) with true lengthscales bigger than $0.3$ can be accurately fitted by only using $10--12$ basis functions.  And, for lengthscales between $0.1$ and $0.3$, $10--22$ basis functions is sufficient.

For a very wiggly $2$D function (let's say with $\ell_1=0.1$ and $\ell_2=0.1$), the approximate number of multivariate basis functions needed is $m*=22\times 22= 484$, which results in significantly faster computation than the regular GP, even with small data sets (i.e., $n<300$) (see Figure~\ref{fig_time_nD_log}). For a very wiggly $3$D function (let's say with $\ell_1=0.1$, $\ell_2=0.1$ and $\ell_3=0.1$), the approximate number of multivariate basis functions needed is $m*=22\times 22\times 22= 10648$, which, unless for small data sets (i.e., $n<1000$), the method is significantly faster than the regular GP. A $3$D function where each dimension has a lengthscale around $0.1$ is not that common in statistical analysis, and thus in many cases the approximation will be significantly faster than excat GP. For $4$D data sets, the method can still be more efficient than the exact GP for moderate-large data sets (i.e., $n>1000$).

Finally, for $D>5$ the method starts to be impractical even for smooth univariate functions and small $n$. However, in these cases the method may still be used for lower dimensional componenets in an additive modeling scheme.
